Brownian motion of main-belt asteroids on human timescales

### 2025-02-24 main.tex
Complete v1

\newcommand{\Var}{\textrm{Var}}
\newcommand{\vcirc}{v_c}
\newcommand{\vrel}{v_{\rm r}}
\newcommand{\Msun}{M_\odot}
\newcommand{\covm}{C}
\newcommand{\lop}{\varpi}
\newcommand{\Lhat}{\hat L}
\usepackage{natbib} 
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{color}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}
%\usepackage{lineno}
%\linenumbers

@@

\begin{abstract}
  The explosion of high-precision astrometric data on main-belt asteroids (MBAs) opens the door to their use for many possible inferences on gravitational and non-gravitational forces present in this region.  We estimate the size of positional shifts on MBAs caused by mutual gravitational encounters with other MBAs that are either omitted from ephemeris models or have uncertain mass estimates. In other words, what is the typical Brownian motion among MBAs that cannot be predicted from the ephemeris, and therefore serves as noise on inferences from the MBAs?  We calculate the RMS azimuthal shift $\sigma_\phi$ of this ``Brownian noise''  by numerical estimation of the distribution of impulse sizes and directions among known MBAs, combined with analytical propagation into future positional uncertainties.  The result depends critically on how well the masses of the larger MBAs are known.  At current level of asteroid-mass knowledge, $\sigma_\phi$ rises to $\approx2$~km or $\approx1$~mas over $T=10$~yr, increasing as $T^{3/2}.$  This is large enough to degrade many inferences from Gaia and LSST MBA data, although the LSST data will also improve MBA mass knowledge enough to lower this Brownian uncertainty by $\ge4\times.$  Radial and vertical Brownian noise  at $T=10$~yr are suppressed by factors $\approx7$ and $\approx45,$ respectively, compared to the azimuthal signal after 10~yr, and grow more slowly with time.  We suggest that for full exploitation of Gaia and LSST MBA data, ephemeris models should include the $\approx1000$~largest asteroids as active bodies with free masses, even if not all are well constrained.  This will correctly propagate the uncertainties from mutual encounters into desired inferences, with the RMS value of positional errors from encounters with less-massive MBAs being limited to $\sigma_\phi\approx60$~m or 30~$\mu$as, small enough to ignore until occultation-based position data become ubiquitously available for MBAs.
\end{abstract}

@@

High-precision measurements of positions of Solar System bodies have
great value to multiple scientific pursuits.  They can reveal
gravitational accelerations due to undiscovered bodies such as a
hypothetical ``Planet X'' \citep[\eg][]{holmanP9,inpopP9,trojans,occultations} or passing primordial
black holes \citep{pbh}, determine
the masses of known bodies \citep[\eg][]{goffin,baer}, test the laws of physics, and characterize 
the non-gravitational forces that drive many forms of planetary
migration, \eg\ the maintainence of the near-Earth asteroid (NEA)
population \citep[reviewed by][]{yarkovsky}.  While most applications of precision orbit determination
have used the 8 major planets as test particles, there are now $>10^6$
known minor planets available for this task.  These small bodies' astrometric
information content is growing extremely rapidly through large-scale
survey projects that can produce
milliarcsecond-scale accuracy for magnitude-limited small-body populations, such as Gaia \citep{gaiass3,gaiafpr},
PanSTARRS,\footnote{\url{https://www2.ifa.hawaii.edu/research/Pan-STARRS.shtml}}
and particularly the upcoming Rubin observatory
\textit{Legacy Survey of Space and Time} (LSST).\footnote{\url{https://rubinobservatory.org/}}  The great majority
of currently-tracked objects are main-belt asteroids (MBAs), so we
wish to investigate the potential power of MBAs as gravitational test
bodies.

@@

While long-term diffusion of orbital elements has been extensively
investigated in the context of planetary-system formation and
planetary migration, we address here the more limited question of how
much the MBAs will have their orbit elements and positions altered
by Brownian motion over a time period $T \lesssim100$~yr of human
observation, in the present dynamical environment.  We will make use
of the currently known population of MBAs, correcting for
incompleteness of current surveys at lower masses---or more precisely,
at fainter absolute magnitudes $H,$ since the $H$ distribution is
known far more accurately than the $M$ distribution.

\section{Calculation overview}
We aim to calculate the Brownian motion variance to an accuracy of a
factor $\approx 2.$
Greater accuracy in the calculation is not warranted
since some of the principal inputs have substantial uncertainty.  The
number of MBAs vs mass is 
uncertain because of unknowns in the conversion
from $H$ to mass; the status of future observations and $\sigma_M$ is
also unclear.  Hence we can determine our quantities only to leading
order in the MBA orbital eccentricity $e$ of the tracer population.
We will assume throughout that the joint
distribution of mass (or $H$) and orbital elements is separable into a
mass distribution and an orbital distribution.  In this scenario, the
Brownian motion is independent of the tracer mass, and involves an
integral over the mass distribution of the deflecting bodies.

We adopt a cylindrical coordinate system $(r,\phi,z),$ with $\zhat$
normal to the initial orbital plane, and $\phi=z=0$ toward the
perhelion of the initial orbit, \ie\ both the initial ascending node
$\Omega_0$ and longitude of perihelion $\lop_0$ equal to zero.  The
unit vectors $\rhat, \phat$ rotate with the target asteroid.  The mean
anomaly at time $t=0$ is $M_0.$

All distances will be given in units of the original semimajor axis $a_0$, and all velocities in units of the circular velocity $\vcirc \equiv \sqrt{G\Msun/a_0}.$  In these units, $G\Msun=1,$ the period of the initial orbit is $2\pi,$ and the (unperturbed) mean anomaly is $M=t+M_0.$

@@

From this knowledge of the impulse distribution, our goal is to obtain
the covariance matrix of the deviations of the target's position from
the initial orbit, after some time $T.$ The position observables are
the range, plus the celestial latitude and longitude of the target.  We will simplify our results by assuming a heliocentric observer, so that the observational position vector is $\vecp\equiv (r,\phi,\theta=z/r).$ The quantity we seek is the covariance matrix $\covm^p$ of the observations attributable to the accumulated gravitational perturbations:
\begin{equation}
  \covm^p  \equiv  \left\langle  \Delta\vecp \Delta\vecp^T\right\rangle,  \label{eq:Cp}
\end{equation}
where the angle brackets indicate an average over possible
realizations of the impulse history, and over the mean anomaly $M$ at the time of observation.  To do so, we will introduce an intermediate set of 6 parameters $\vecq$ describing the deviations of the orbital elements induced by the impulses.  The $\vecq$ components will be selected to be equal to zero before any impulses are applied, and each responds linearly to $\vecI$ at $|I|\ll 1.$
We will derive the matrix $\matA$ that describes the orbital-element shifts at time $T$ that arise from an impulse at time $t_i$: 
\begin{equation}
  \Delta \vecq(e_0,T,t_i) = \matA(e_0, T, t_i) \vecI.
  \label{eq:A}
\end{equation}
In our first-order perturbation theory, $\vecq(T)$ will be the sum of the $\Delta\vecq_i$ imparted by all impulses applied at times $0<t_i<T.$  Because the impulses are uncorrelated, $\vecq$ will therefore be the result of a random walk.  The distribution will have a covariance matrix $\covm^q(T) \equiv \left\langle \vecq(T) \vecq^T(T) \right\rangle$ whose elements are
\begin{eqnarray}
  \covm^q_{jk}(e_o,T) & = & \left\langle \sum_{i,\gamma} \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) I^2_{i,\gamma} \right\rangle \\
           & = & \sum_\gamma \int dt_i \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) \left\langle n I_\gamma^2\right\rangle.
\label{eq:Cqjk}
\end{eqnarray}
In the first line, the sum $i$ runs over the impulses and $\gamma$ runs over the components $r,\phi,z$ of the impulse. The second line evaluates the expectation value of averaging over realizations of the random walk of impulses, exploiting the independence of the individual impulses from each other.

The last element of our calculation will be a conversion from the
element shifts $\vecq$ into the observed displacements $\Delta\vecp$
at time $T.$  In linear perturbation theory this will again be expressible as a matrix
\begin{eqnarray}
\label{eq:B}
  \Delta\vecp(e_0,T) & = & \matB(e_0,T) \vecq(T) \\
  \Rightarrow \quad 
  \covm^p_{\alpha\beta}(e_0,T) & = & \sum_{jk} B_{\alpha j}(e_0,T) B_{\beta k}(e_0,T) \covm^q_{jk}(T) 
\label{eq:covp}
\end{eqnarray}
which we would wish to average over the phase of the MBA orbit or mean anomaly $M$ at the time of observation $T.$

Section~\ref{sec:impulse} describes the estimation of $\left\langle
  nI_\alpha^2\right\rangle$ for MBA regions.
Section~\ref{sec:propagation} derives the forms of $\matA(e_0,T,t_i),$
and $\matB(e_0,T).$  Section~\ref{sec:results} combines these and
summarizes the results and their implications.

\section{Impulse distribution}
\label{sec:impulse}
\subsection{Derivation}
Under the assumption that the distributions of masses and orbital
elements of MBAs are separable, the quantities needed to describe the
random walk of the asteroid's orbits can be expressed as an integral
over the absolute magnitude $H$ of the deflectors, the velocity $v$,
impact parameter $b,$ and polar/azimuthal angles $\theta,\phi$ of the
unit vector of the impulse direction $\bhat$:
\begin{eqnarray}
  \left\langle n I_\gamma^2 \right\rangle & = & \int
                                              dH\,dv\,db\,d\theta\,d\phi
                                             \frac{dN}{dH} \frac{dn}{dv\,db\,d\theta\,d\phi}
                                              I^2_\gamma(H,v,b,\theta\phi)
  \\
  & = & \int dH \frac{dN}{dH} \int dv
        \frac{dn}{dA\,dv\,d\theta\,d\phi} \hat b_\gamma^2(\theta,\phi)
        \int_{b_{\rm min}}^{b_{\rm max}} 2\pi b\,db\,
        \left(\frac{2\sigma_M(H)}{bv}\right)^2 \\
  & = & 8\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int dv
        \frac{dn}{dA\,dv} v^{-2} \left\langle\hat b_\gamma^2(v)\right\rangle
        \log(b_{\rm max}/b_{\rm min}).
\end{eqnarray}
The second row adopts the (numerically verified) assumption that the
distribution of impact parameter $b$ will always be $\propto dA=2\pi
b\,db$ in the range of $b$ of interest to us.  The mean rate of encounters
between a deflector-tracer pair, as a function of the impact velocity
and geometry, is $dn/dA\,dv\,d\theta\,d\phi,$ which we will determine
through numerical measurements with the orbits of all known MBAs. The
$H$ distribution of MBAs is $dN/dH.$ We denote as $\sigma_M(H)$
the RMS error in the mass of MBAs at $H$ from the value used in the ephemeris model.
We use $\sigma_M$ rather than $M$ since we are seeking the size of
deviations from the ephemeris model.  For MBAs that are not included
in the ephemeris model, $\sigma_M=M(H),$ the full mass of the deflector.

@@

For $b_{\rm min},$ we adopt the criterion that encounters at
sufficiently small $b$ will generate sufficiently large impulses $I$
that observations of the tracer will detect this individual impulse at
levels well above measurement noise.  This would mean that an
ephemeris model could include the mass of the deflector asteroid, and
this mass could be determined from fitting the data of the tracer.
Such encounters would thus no longer be considered source of
stochastic Brownian motion.  Finding all such encounters among known
MBAs is feasible, and in fact is forecasted for the next decade's LSST
observations by
\citet{negin}. The fitting of the tracer data for the mass of the
deflector is also a small computational effort.

If we crudely choose some threshold
$I_{\rm det}$ of impulse as being large enough to generate high-$S/N$
deflections on its tracer, then our condition becomes $2M(H)/bv <
I_{\rm det},$ or $b_{\rm min}=2M(H)/vI_{\rm det}.$  We then have
\begin{equation}
   \left\langle n I_\gamma^2 \right\rangle = 
2\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int_{\sqrt{2M(H)/I_{\rm
      det}}}^\infty \frac{dv}{v^2} 
        \frac{dn}{dA\,dv}  \left\langle\hat b_\gamma^2(v)\right\rangle
        \log \left[v^2 I_{\rm det}/2M(H)\right].
        \label{eq:nI2}
      \end{equation}
The lower bound on the $v$ integral marks the speed below which
$b_{\rm max}<b_{\rm min}.$  Since $b_{\rm min}$ and $b_{\rm max}$
appear logarithmically, the estimated diffusion is relatively insensitive to choices of
their values.

@@

\subsubsection{Encounter rates and geometries}
The pairwise interaction rate $dn/dA\,dv$ and the impulse direction
distributions $\langle \hat b^2_\gamma \rangle$ are estimated from a
numerical integration of all of the MBA orbits available
from the Minor Planet Center (MPC) as of 6 January 2025.   We retain as
potential deflectors the 1.26~million bodies with $1.8<a<4.2$~AU,
observations at multiple oppositions, and uncertainty values $U\le5.$
We designate each source as being an ``Inner'' MBA, with semi-major axes between the 4:1 and 3:1
mean-motion resonances of Jupiter; ``Middle'' MBA between the 3:1 and
5:2 resonances; ``Outer'' MBAs between the 5:2 and 2:1
resonances; the remaining ``Other'' objects are a small minority that
contribute very little to the total diffusion rate, and we will ignore.  Our working
assumption will be that the MPC objects are an unbiased sample of the
orbital-element distribution within each of the Inner, Middle, and
Outer belt regions.

@@

For each pair of tracer-deflector regions (\eg\ Inner-Inner), we combine the list of
events with the counts of candidate tracers and deflectors to
calculate $dn/dA\,dv,$ the event rates per eligible pair; and the mean
geometry of the encounter, $\langle \hat b^2_\gamma \rangle,$ as a
function of $v$, all plotted in Figure~\ref{fig:rtz}.
Here it is apparent that the $\hat b_\phi$ component is
larger than $\hat b_z$ and $\hat b_r,$ but there is significant
variation with MBA class and with $v$.  An important result is that
the quantity $v^{-2} (dN/dA\,dv)$ is always well-behaved as $v\rightarrow 0,$
so we do not need any special treatment to sum the integrals in
\eqq{eq:nI2}.  The $v$ integrand peak is in the range $0.03<v/v_c<0.1.$

@@

To give some context, a
comprehensive investigation of MBA masses from mutual scattering
events by \citet{goffin} reports typical $\sigma_M\approx
5\times10^{-13}$ for 30 MBAs. But we can expect substantial improvement
in the LSST era when a millions of potential tracer asteroids attain
mas-level astrometry. \citet{negin} use a simulated LSST sequence of observations to estimate the impulse (and mass) uncertainties from analyzing LSST's data on mutual encounters of the known asteroids. The tracer MBAs for those events are essentially an ubiased selection from the known MBAs, and the left-hand plot of Figure~\ref{fig:sigmaI} shows the distribution of the uncertainty $\sigma_I$ derived from each tracer's LSST data after marginalizing over its initial state vector.  This histogram marginalizes over the apparent magnitude and observing conditions of the tracer, the date of the encounter, and the direction $\bhat$ of the impulse.  The right-hand plot of Figure~\ref{fig:sigmaI} forecasts the distribution of $\sigma_M$ that we can expect for deflectors.  This forecast is made as follows: we treat each of the 5000 tracer MBAs that we have selected from each region as deflectors, since we have a list of all of their encounters with known MBAs with $b<0.03$~AU.  For each such encounter, we select at random one $\sigma_I$ from the distribution found by \citet{negin} (Figure~\ref{fig:sigmaI}, left), and calculate $\sigma_M=(bv/2)\sigma_I.$  For each deflector, we then calculate (a) the individual encounter with lowest $\sigma_M$; (b) the combined constraint $\left(\sum \sigma_{M,i}^2\right)^{-1/2}$ from the 10 lowest $\sigma_M$ encounters for that deflector; and (c) the combined constraint for all $b<0.03$~AU encounters (typically $\sim1000$ of them).  The right-hand plot of Figure~\ref{fig:sigmaI} gives the histograms of these three measures of $\sigma_M.$  We see that most of the available information on any given deflector's mass comes from its 10 most-informative encounters.  It would be computationally very easy to solve for the masses of $O(10^4)$ MBAs by including the deflections they impart on $O(10^5)$ other MBAs in an ephemeris fit.
The RMS $\sigma_M$ using LSST data from the 10 most-informative tracer encounters with a given deflector is estimated as $\sigma_M=2.6\times10^{-14}M_\odot.$ There are $\approx450$ MBAs with mass above this forecasted RMS $\sigma_M.$  This is a conservative estimate in that we ignore any observational information before LSST, and we also do not consider the information increase that will come from tracking the $\approx10$-fold increase in number of known MBAs that LSST will enable.

>>>>

\newcommand{\Var}{\textrm{Var}}
\newcommand{\vcirc}{v_c}
\newcommand{\vrel}{v_{\rm r}}
\newcommand{\Msun}{M_\odot}
\newcommand{\covm}{C}
\newcommand{\lop}{\varpi}   % longitude of perihelion
\newcommand{\ma}{\ell}  % Mean anomaly

\newcommand{\Lhat}{\hat L}
\usepackage{natbib} 
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{color}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}
%\usepackage{lineno}
%\linenumbers

@@

\begin{abstract}
  The explosion of high-precision astrometric data on main-belt
  asteroids (MBAs) opens the door to their use for many possible
  inferences on gravitational and non-gravitational forces present in
  this region.  We estimate the size of positional shifts on MBAs
  caused by mutual gravitational encounters with other MBAs that are
  either omitted from ephemeris models or have uncertain mass
  estimates. In other words, what is the typical Brownian motion among
  MBAs that cannot be predicted from the ephemeris, and therefore
  serves as noise on inferences from the MBAs?  We calculate the RMS
  azimuthal shift $\sigma_\phi$ of this ``Brownian noise''  by
  numerical estimation of the distribution of impulse sizes and
  directions among known MBAs, combined with analytical propagation
  into future positional uncertainties.  The result depends strongly
  on how well the masses of the larger MBAs are known.  At current
  level of asteroid-mass knowledge, $\sigma_\phi$ rises to
  $\approx2$~km or $\approx1$~mas over $T=10$~yr, increasing as
  $T^{3/2}.$  This is large enough to degrade many inferences from
  Gaia and LSST MBA data, although the LSST data will also improve MBA
  mass knowledge enough to lower this Brownian uncertainty by
  $\ge4\times.$  Radial and vertical Brownian noise  at $T=10$~yr are
  suppressed by factors $\approx7$ and $\approx45,$ respectively,
  compared to the azimuthal signal after 10~yr, and grow at $T^{3/2}$
  and $T^{1/2}.$  We suggest that for full exploitation of Gaia and
  LSST MBA data, ephemeris models should include the
  $\approx1000$~largest asteroids as active bodies with free masses,
  even if not all are well constrained.  This will correctly propagate
  the uncertainties from these 1000 sources' deflections into desired
  inferences. The RMS value of deflections from less-massive MBAs is limited to $\sigma_\phi\approx60$~m or 30~$\mu$as, small enough to ignore until occultation-based position data become ubiquitously available for MBAs.
\end{abstract}

@@

High-precision measurements of positions of Solar System bodies have
great value to multiple scientific pursuits.  They can reveal
gravitational accelerations due to undiscovered bodies such as a
hypothetical ``Planet X'' \citep[\eg][]{holmanP9,inpopP9,trojans,occultations} or passing primordial
black holes \citep[\eg][]{pbh}, determine
the masses of known bodies \citep[\eg][]{goffin,baer}, test the laws
of physics \citep[\eg][]{inpopgraviton}, and characterize 
the non-gravitational forces that drive many forms of planetary
migration, \eg\ the maintainence of the near-Earth asteroid (NEA)
population \citep[reviewed by][]{yarkovsky}.  While most applications of precision orbit determination
have used the 8 major planets as test particles, there are now $>10^6$
known minor planets available for this task.  These small bodies' astrometric
information content is growing extremely rapidly through large-scale
survey projects that can produce
milliarcsecond-scale accuracy for magnitude-limited small-body populations, such as Gaia \citep{gaiass3,gaiafpr},
PanSTARRS,\footnote{\url{https://www2.ifa.hawaii.edu/research/Pan-STARRS.shtml}}
and particularly the upcoming Rubin observatory
\textit{Legacy Survey of Space and Time} (LSST).\footnote{\url{https://rubinobservatory.org/}}  The great majority
of currently-tracked objects are main-belt asteroids (MBAs), so we
wish to investigate the potential power of MBAs as gravitational test
bodies.

@@

While long-term diffusion of orbital elements has been extensively
investigated in the context of planetary-system formation and
planetary migration, we address here the more limited question of how
much the MBAs will have their orbit elements and positions altered
by Brownian motion over a time period $T \lesssim100$~yr of human
observation, in the present dynamical environment.  Indeed essentially
all of the relevant deflector bodies and circumstances of encounters
are known.  We will make use
of the currently known population of MBAs, correcting for
incompleteness of current surveys at lower masses---or more precisely,
at fainter absolute magnitudes $H,$ since the $H$ distribution is
known far more accurately than the $M$ distribution.

\section{Calculation overview}
We aim to calculate the Brownian motion variance to an accuracy of a
factor $\approx 2.$
Greater accuracy in the calculation is not warranted
since some of the principal inputs have substantial uncertainty.  The
number of MBAs vs mass is 
uncertain because of unknowns in the conversion
from $H$ to mass; the status of future observations and $\sigma_M$ is
also unclear.  We are hence justified in retaining only the
contributions to our results that are leading
order in the MBA orbital eccentricity $e$ of the tracer population.
We will assume throughout that the joint
distribution of mass (or $H$) and orbital elements is separable into a
mass distribution and an orbital distribution.  In this scenario, the
Brownian motion is independent of the tracer mass, and involves an
integral over the mass distribution of the deflecting bodies.

We adopt a cylindrical coordinate system $(r,\phi,z),$ with $\zhat$
normal to the initial orbital plane, and $\phi=z=0$ toward the
perhelion of the initial orbit, \ie\ both the initial ascending node
$\Omega_0$ and longitude of perihelion $\lop_0$ equal to zero.  The
unit vectors $\rhat, \phat$ rotate with the target asteroid.  The mean
anomaly at time $t=0$ is $\ma_0.$

All distances will be given in units of the original semimajor axis $a_0$, and all velocities in units of the circular velocity $\vcirc \equiv \sqrt{G\Msun/a_0}.$  In these units, $G\Msun=1,$ the period of the initial orbit is $2\pi,$ and the (unperturbed) mean anomaly is $\ma=t+\ma_0.$

@@

From this knowledge of the impulse distribution, our goal is to obtain
the covariance matrix of the deviations of the target's position from
the initial orbit, after some time $T.$ The position observables are
the range, plus the celestial latitude and longitude of the target.  We will simplify our results by assuming a heliocentric observer, so that the observational position vector is $\vecp\equiv (r,\phi,\theta=z/r).$ The quantity we seek is the covariance matrix $\covm^p$ of the observations attributable to the accumulated gravitational perturbations:
\begin{equation}
  \covm^p  \equiv  \left\langle  \Delta\vecp \Delta\vecp^T\right\rangle,  \label{eq:Cp}
\end{equation}
where the angle brackets indicate an average over possible
realizations of the impulse history, and over the mean anomaly $\ma$ at the time of observation.  To do so, we will introduce an intermediate set of 6 parameters $\vecq$ describing the orbital elements induced by the impulses.  The $\vecq$ components will be selected to respond linearly to $\vecI$ at $|I|\ll 1.$
We will derive the matrix $\matA$ that describes the orbital-element shifts at time $T$ that arise from an impulse at time $t_i$: 
\begin{equation}
  \Delta \vecq(e_0,T,t_i) = \matA(e_0, T, t_i) \, \vecI.
  \label{eq:A}
\end{equation}
In our first-order perturbation theory, $\Delta\vecq(T)$ will be the sum of the $\Delta\vecq_i$ imparted by all impulses applied at times $0<t_i<T.$  Because the impulses are uncorrelated, $\vecq$ will therefore be the result of a random walk.  The distribution will have a covariance matrix $\covm^q(T) \equiv \left\langle \Delta\vecq(T) \Delta\vecq^T(T) \right\rangle$ whose elements are
\begin{eqnarray}
  \covm^q_{jk}(e_o,T) & = & \left\langle \sum_{i,\gamma} \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) I^2_{i,\gamma} \right\rangle \\
           & = & \sum_\gamma \int dt_i \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) \left\langle n I_\gamma^2\right\rangle.
\label{eq:Cqjk}
\end{eqnarray}
In the first line, the sum $i$ runs over the impulses and $\gamma$ runs over the components $r,\phi,z$ of the impulse. The second line evaluates the expectation value of averaging over realizations of the random walk of impulses, exploiting the independence of the individual impulses from each other.

The last element of our calculation will be a conversion from the
element shifts $\Delta\vecq$ into the observed displacements $\Delta\vecp$
at time $T.$  In linear perturbation theory this will again be expressible as a matrix
\begin{eqnarray}
\label{eq:B}
  \Delta\vecp(e_0,T) & = & \matB(e_0,T) \vecq(T) \\
  \Rightarrow \quad 
  \covm^p_{\alpha\beta}(e_0,T) & = & \sum_{jk} B_{\alpha j}(e_0,T)
                                     B_{\beta k}(e_0,T) \, \covm^q_{jk}(T) 
\label{eq:covp}
\end{eqnarray}
which we would wish to average over the phase of the MBA orbit or mean anomaly $\ma$ at the time of observation $T.$

Section~\ref{sec:impulse} describes the estimation of $\left\langle
  nI_\alpha^2\right\rangle$ for MBA regions.
Section~\ref{sec:propagation} derives the forms of $\matA(e_0,T,t_i),$
and $\matB(e_0,T).$  Section~\ref{sec:results} combines these and
summarizes the results.  The implications of these results for use of
MBA observations for high-precision inferences are discussed in Section~\ref{sec:discuss}.

\section{Impulse distribution}
\label{sec:impulse}
\subsection{Derivation}
Under the assumption that the distributions of masses and orbital
elements of MBAs are separable, the quantities needed to describe the
random walk of the asteroid's orbits can be expressed as an integral
over the absolute magnitude $H$ of the deflectors, the velocity $v$,
impact parameter $b,$ and polar/azimuthal angles $\theta,\phi$ of the
unit vector of the impulse direction $\bhat$:
\begin{eqnarray}
  \left\langle n I_\gamma^2 \right\rangle & = & \int
                                              dH\,dv\,db\,d\theta\,d\phi
                                             \frac{dN}{dH} \frac{dn}{dv\,db\,d\theta\,d\phi}
                                              I^2_\gamma(H,v,b,\theta\phi)
  \\
  & = & \int dH \frac{dN}{dH} \int dv
        \frac{dn}{dA\,dv\,d\theta\,d\phi} \hat b_\gamma^2(\theta,\phi)
        \int_{b_{\rm min}}^{b_{\rm max}} 2\pi b\,db\,
        \left(\frac{2\sigma_M(H)}{bv}\right)^2 \\
  & = & 8\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int dv
        \frac{dn}{dA\,dv} v^{-2} \left\langle\hat b_\gamma^2(v)\right\rangle
        \log(b_{\rm max}/b_{\rm min}).
\end{eqnarray}
The second row adopts the (numerically verified) assumption that the
distribution of impact parameter $b$ will always be $\propto dA=2\pi
b\,db$ in the range of $b$ of interest to us.  The mean rate of encounters
between a deflector-tracer pair, as a function of the impact velocity
and geometry, is $dn/dA\,dv\,d\theta\,d\phi,$ which we will determine
through numerical measurements with the orbits of all known MBAs. The
$H$ distribution of MBAs is $dN/dH.$ We define $\sigma_M(H)$ via
\begin{equation}
  \sigma_M(H) \equiv \left\{\begin{array}{cl}
                              \sigma_M, & M(H) > \sigma_M \\
                              M(H),        & M(H) \le \sigma_M
                            \end{array} \right.
  \label{eq:sigmaMH}
\end{equation}
such that it represents the error in the ephemeris due to uncertainty
in the masses of MBAs included in the ephemeris (first row), and due
to the entirety of the masses of MBAs omitted from the ephemeris
(second row).

@@

For $b_{\rm min},$ we adopt the criterion that encounters at
sufficiently small $b$ will generate sufficiently large impulses $I$
that observations of the tracer will detect this individual impulse at
levels well above measurement noise.  This would mean that an
ephemeris model could include the mass of the deflector asteroid, and
this mass could be usefully determined from fitting the data of just this
single tracer.
Such encounters would thus no longer be considered source of
stochastic Brownian motion.  Finding all such encounters among known
MBAs is feasible, and in fact is forecasted for the next decade's LSST
observations by
\citet{negin}.

If we crudely choose some threshold
$I_{\rm det}$ of impulse as being large enough to generate high-$S/N$
deflections on its tracer, then our condition becomes $2M(H)/bv <
I_{\rm det},$ or $b_{\rm min}=2M(H)/vI_{\rm det}.$  We then have
\begin{equation}
   \left\langle n I_\gamma^2 \right\rangle = 
8\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int_{\sqrt{2M(H)/I_{\rm
      det}}}^\infty \frac{dv}{v^2} 
        \frac{dn}{dA\,dv}  \left\langle\hat b_\gamma^2(v)\right\rangle
        \log \left[v^2 I_{\rm det}/2M(H)\right].
        \label{eq:nI2}
      \end{equation}
The lower bound on the $v$ integral marks the speed below which
$b_{\rm max}<b_{\rm min}.$  Since $b_{\rm min}$ and $b_{\rm max}$
appear logarithmically, the estimated diffusion is relatively insensitive to choices of
their values.

@@

\subsubsection{Encounter rates and geometries}
The pairwise interaction rate $dn/dA\,dv$ and the impulse direction
components $\langle \hat b^2_\gamma \rangle$ are estimated from a
numerical integration of all of the MBA orbits available
from the Minor Planet Center (MPC) as of 6 January 2025.   We retain as
potential deflectors the 1.26~million bodies with $1.8<a<4.2$~AU,
observations at multiple oppositions, and uncertainty values $U\le5.$
We designate each source as being an ``Inner'' MBA, with semi-major axes between the 4:1 and 3:1
mean-motion resonances of Jupiter; ``Middle'' MBA between the 3:1 and
5:2 resonances; ``Outer'' MBAs between the 5:2 and 2:1
resonances; the remaining ``Other'' objects are a small minority that
contribute very little to the total diffusion rate, and we will ignore.  Our working
assumption will be that the MPC objects are an unbiased sample of the
orbital-element distribution within each of the Inner, Middle, and
Outer belt regions.

@@

For each pair of tracer-deflector regions (\eg\ Inner-Inner), we combine the list of
events with the counts of candidate tracers and deflectors to
calculate $dn/dA\,dv,$ the event rates per eligible pair; and the mean
geometry of the encounter, $\langle \hat b^2_\gamma \rangle,$ as a
function of $v$, all plotted in Figure~\ref{fig:rtz}.
Here it is apparent that the $\hat b_\phi$ component is, on average,
larger than $\hat b_z$ and $\hat b_r,$ but there is significant
variation with MBA class and with $v$.  An important result is that
the quantity $v^{-2} (dN/dA\,dv)$ is always well-behaved as $v\rightarrow 0,$
so we do not need any special treatment to sum the integrals in
\eqq{eq:nI2}.  The $v$ integrand peak is in the range $0.03<v/v_c<0.1.$

@@

To give some context, a
comprehensive investigation of MBA masses from mutual scattering
events by \citet{goffin} reports typical $\sigma_M\approx
5\times10^{-13}$ for 30 MBAs (although the list is biased toward
better-measured events). But we can expect substantial improvement
in the LSST era when a millions of potential tracer asteroids attain
mas-level astrometry. \citet{negin} use a simulated LSST sequence of
observations to estimate the impulse (and mass) uncertainties from
analyzing LSST's data on mutual encounters of the known asteroids. The
tracer MBAs for those events are essentially an unbiased selection
from the known MBAs, and the left-hand plot of Figure~\ref{fig:sigmaI}
shows the distribution of the uncertainty $\sigma_I$ derived from each
tracer's LSST data after marginalizing over its initial state vector.
This histogram marginalizes over the apparent magnitude and observing
conditions of the tracer, the date of the encounter, and the direction
$\bhat$ of the impulse.  The right-hand plot of
Figure~\ref{fig:sigmaI} forecasts the distribution of $\sigma_M$ that
we can expect for deflectors.  This forecast is made as follows: we
treat each of the 5000 tracer MBAs that we have selected from each
region as deflectors, since we have a list of all of their encounters
with known MBAs with $b<0.03$~AU.  For each such encounter, we select
at random one $\sigma_I$ from the distribution for LSST data on known
MBAs forecasted by \citet{negin} (Figure~\ref{fig:sigmaI}, left), and
calculate $\sigma_M=(bv/2)\sigma_I.$  For each deflector, we then
calculate (a) the individual encounter with lowest $\sigma_M$; (b) the
combined constraint $\left(\sum \sigma_{M,i}^2\right)^{-1/2}$ from the
10 lowest $\sigma_M$ encounters for that deflector; and (c) the
combined constraint for all $b<0.03$~AU encounters (typically
$\sim1000$ of them).  The right-hand plot of Figure~\ref{fig:sigmaI}
gives the histograms of these three measures of $\sigma_M.$  We see
that most of the available information on any given deflector's mass
comes from its 10 most-informative encounters.  It should not be a
computational barrier to solve for the masses of $O(10^4)$ MBAs by including the deflections they impart on $O(10^5)$ other MBAs in an ephemeris fit.
The RMS $\sigma_M$ using LSST data from the 10 most-informative tracer encounters with a given deflector is estimated as $\sigma_M=2.6\times10^{-14}M_\odot.$ There are $\approx450$ MBAs with mass above this forecasted RMS $\sigma_M.$  This is a conservative estimate in that we ignore any observational information before LSST, and we also do not consider the information increase that will come from tracking the $\approx10$-fold increase in number of known MBAs that LSST will enable.
